---
title: "Mono Lake Visualization"
description: |
  This was completed as part of an assignment for Allison Horst's ESM 206 course.
author:
  - name: Craig Kopulsky
    url: {}
date: 03-14-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

This is a visualization of the annual water level change at Mono Lake, in Lee Vining, California, from 1850-2017.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(ggalt)
library(ggrepel)
```

```{r, include=FALSE}
mono_lake_data <- read.csv(here("_posts", "2021-03-14-mono-lake-visualization","data","mono_lake_annual_levels.csv"), skip = 5) %>%  # reading in data
  clean_names() %>% # converting column names to lowercase_snakecase
  mutate(key_event = case_when( # creating a variable to hold two labels that will be added to chart later
    year == 1941 ~ "1941 - Start of LADWP water diversion",
    year == 1994 ~ "1994 - Decision 1631 sets \nlake level of 6392 ft \nto protect Mono Lake"
  ))
```



```{r, include=FALSE}
spline_change <- as.data.frame(spline(mono_lake_data$year, mono_lake_data$lake_level_feet_above_sea_level)) # using spline() to create a data set that creates a curved line that will go through all data points on following chart
```


```{r, echo = FALSE, fig.width = 10}
mono_lake_plot <- ggplot(data = mono_lake_data, aes(x = year, y = lake_level_feet_above_sea_level)) + # greating gg plot with mono lake data
  geom_line(data = spline_change, aes(x = x, y = y), size = 1, color = "cyan4") + # adding curved line
  geom_ribbon(data = mono_lake_data, aes(ymin = -Inf, ymax = lake_level_feet_above_sea_level), fill = "cadetblue2", alpha = .2) + # adding ribbon beneath line
  geom_hline(yintercept = 6360, color = "coral2", size = 1) + # adding hline
  annotate("text", x = 2017, y = 6360,
           label = "   Negative impacts \n   from salinity",
           hjust = 0, size = 2.5, fontface = "bold") + # adding annotation
  geom_hline(yintercept = 6377, color = "coral", size = 1) + # adding hline
  annotate("text", x = 2017, y = 6377,
           label = "   Land bridge formation \n   leads to predation",
           hjust = 0, size = 2.5, fontface = "bold") + # adding annotation
  geom_hline(yintercept = 6350, color = "coral3", size = 1) + # adding hline
  annotate("text", x = 2017, y = 6350,
           label = "   Acute impacts from \n   salinity changes",
           hjust = 0, size = 2.5, fontface = "bold") + # adding annotation
  geom_label_repel(aes(label = key_event),
                   fill = "aliceblue",
                   size = 3,
                   nudge_y = 25) + # adding key event labels
  scale_x_continuous(breaks = seq(1850,2017,20), expand = c(0,0)) + # adjusting x axis
  ylim(6350,6450) + # adjusting y axis
  labs(x = "Year",
       y = "Lake Level (feet above sea level)") + # changing labels
  theme_minimal() # theme minimal

 mono_lake_plot + # creating some white space to right of chart for annotations to show
  theme(plot.margin = unit(c(0.5,6,0.5,0.5), "lines")) +
  coord_cartesian(clip = "off")
```

**Figure 1:** Mono Lake water level annual change (1850 - 2017). Key events impacting the lake as well as key lake water levels that would have a major impact on lake ecology are noted. Data: Mono Basin Clearinghouse.


**Citation:**

Mono Basin Clearinghouse, Mono Lake Committee. "Mono Lake Levels 1850-present". https://www.monobasinresearch.org/data/levelyearly.php. (October 2017).




