---
title: "California Oil Spills - 2008"
description: |
  This was completed as part of an assignment for Allison Horst's ESM 244 course.
author:
  - name: Craig Kopulsky
    url: {}
date: 02-21-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(tmap)
library(gstat)
library(stars)
library(ggiraph)

```

## Introduction

This analysis visualizes California oil spill data in 2008 from the California Department of Fish and Game, Office of Spill Prevention and Response.

```{r, message=FALSE, warning=FALSE, results='hide', echo=TRUE}
# Read in the oil spill data
oil_spill_data <- st_read(here("_posts/2021-02-21-california-oil-spills-2008/oil_spill_data"), layer = "Oil_Spill_Incident_Tracking_%5Bds394%5D") %>% 
  clean_names()

# Check the projection
st_crs(oil_spill_data)

# Read in the CA county data (TIGER shapefile):
ca_counties <- read_sf(here("_posts/2021-02-21-california-oil-spills-2008/ca_counties"), layer = "CA_Counties_TIGER2016") %>% 
  clean_names() %>% 
  select(name)

# Check the projection
st_crs(ca_counties)
ca_counties <- st_transform(ca_counties, st_crs(oil_spill_data))
```


```{r, include = FALSE}
# Make a quick ggplot:
ggplot() +
  geom_sf(data = ca_counties) +
  geom_sf(data = oil_spill_data, aes(color = specificlo))
```



## Visualizing California oil spill locations and spill environment 

```{r, message=FALSE, warning=FALSE}
tmap_mode("view")

tm_shape(oil_spill_data) +
  tm_dots("specificlo", id = "oesnumber", title = "Spill Environment")
```

**Figure 1:** Interactive map of California showing the location of oil spill events and the type of environment on which those events occurred in 2008. Number shown upon hover over each spill is the unique OES Number, a unique control number assigned by Cal EMA for tracking spills.

## Visualizing oil spills per county

```{r, message=FALSE, warning=FALSE}
# Wrangling ca_counties, alining variable name for county name to match oil_spill_data variable
ca_counties_spillcount <- ca_counties %>%
  rename(localecoun = name)
```

```{r, message=FALSE, warning=FALSE}
# Wrangling oil_spill_data, filtering out marine spills, counting number of spills per county
oil_spill_data_spillcount <- oil_spill_data %>% 
  filter(inlandmari == "Inland") %>% 
  count(localecoun)
```

```{r, message=FALSE, warning=FALSE}
# Joining the two datafiles into one that contains the spill counts by county in CA
spill_count_by_county <- ca_counties_spillcount %>% 
  st_join(oil_spill_data_spillcount)
```

```{r, message=FALSE, warning=FALSE}
# Creating a cloropleth plot using the number of oil spills by county as the fill color
ggplot(data = spill_count_by_county) +
  geom_sf(aes(fill = n), color = "black", size = .1) +
  scale_fill_gradientn(colors = c("lightgrey","burlywood3","burlywood4","black"), na.value = "white") +
  theme_minimal() +
  labs(fill = "Number of inland oil spills")
```

**Figure 2:** Cloropleth plot of California showing the number of spills per county in 2008. Only inland spills are included in plot. Counties with missing spill data are shown as white.


## Citation

California Department of Fish and Game, Office of Spill Prevention and Response. "Oil Spill Incident Tracking [ds394]" [digital map]. "California State Geoportal". 2009-07-23
. https://gis.data.ca.gov/datasets/7464e3d6f4924b50ad06e5a553d71086_0/data. (2021-02-20).

