---
title: "Cetacean Density off the Coast of California"
description: |
  This was completed as part of an assignment for Allison Horst's ESM 244 course.
author:
  - name: Craig Kopulsky
    url: {}
date: 03-14-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(raster) ### NOTE: attaching this BEFORE tidyverse
library(tidyverse)
library(janitor)
library(here)
library(sf)
library(fasterize)
library(rnaturalearth)
```

**Introduction:**

This report analyzes spatial data for the probability of occurrence of 36 different cetacean species off the coast of California. A probability of .6 and above is used as a threshold to determine species presence at any point on the map for each of these species.


```{r, message=FALSE, warning=FALSE}
# reading in cetacean raster data and stacking data into one data frame
cetaceans_file <- list.files(path = here('_posts/2021-03-14-cetacean-density-off-the-coast-of-california/data/ca_cetaceans'), full.names = TRUE, pattern = ".tif")

ca_cetaceans <- stack(cetaceans_file)
```


```{r, message=FALSE, warning=FALSE}
# creating a function to set the threshold for a cetacean species to be counted as present in a given location at .6 and above
is_cetacean_pres <- function(x, thresh = .6) {
  y <- ifelse(x >= thresh, 1, 0)
  return(y)
}

# using the above function on the ca_cetaceans dataset to convert cetacean probabilities at each point to 1 for present and 0 for not present
cetacean_pres <- calc(ca_cetaceans, fun = is_cetacean_pres)
#plot(cetacean_pres, col = c('red', 'green4'))
```

**Data Visualization:**


```{r, message=FALSE, warning=FALSE}
# turning cetacean_pres raster into a data frame
cetacean_pres_df <- raster::rasterToPoints(cetacean_pres) %>%
  as.data.frame()
```


```{r, message=FALSE, warning=FALSE}
# creating a data frame that contains the sum of how many cetacean species were present at any given point in the raster
cetacean_density <- calc(cetacean_pres, fun = sum, na.rm = TRUE)

plot(cetacean_density)
```

**Figure 1:** Species richness raster map of 36 cetacean species off the coast of California. Color indicates the number of species present at a given point.

```{r, fig.width = 10, fig.height= 10, message=FALSE, warning=FALSE, results='hide'}
# downloading shapefiles of north america
coast <- ne_download(scale = 110, type = 'land', category = 'physical', returnclass = "sf")
states <- ne_download(scale = 110, type = 'states', category = 'cultural', returnclass = "sf")

# filtering shape file for california
ca <- states %>% 
  filter(name == "California")

# turning raster cetacean_density into a data frame
cetacean_density_df <- raster::rasterToPoints(cetacean_density) %>%
  as.data.frame()

#creating static map using density data frame and overlaying north america and california shapes
ggplot() +
  geom_raster(data = cetacean_density_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_distiller(type = "seq", palette = "BuPu", aesthetics = "fill", direction = 1) +
  geom_sf(data = coast, fill = "burlywood4") +
  geom_sf(data = ca, fill = "burlywood4") +
  coord_sf(xlim = c(-125, -115), ylim = c(32,38), expand = 0) +
  labs(x = "", y = "", fill = "Cetacean Richness")
  
```  

**Figure 2:** Static map depicting species precense for 36 cetacean species off the coast of California. Color indicates the number of species present at a given point.


Citation:

Kaschner, K., Rius-Barile, J., Kesner-Reyes, K., Garilao, C., Kullander, S., Rees, T., & Froese, R. (2016). AquaMaps: Predicted range maps for aquatic species. www.aquamaps.org






